\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}

\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\var}{var}

\begin{document}

\section{Description of the problem: correlated evolution}

Suppose we have measured trait values in a number of individuals from each of several taxa related by a known phylogenetic tree.
It is hypothesized that the evolutionary dynamics of some of these traits
are affected by the values of some other traits,
thus introducing correlation between them.
The driving trait(s) may be observed or unobserved.
The goal is to test this hypothesis,
which we will aim to do by estimating the strength of that correlation
in a Bayesian framework.
Since we have measured multiple individuals within each species,
we have two sorts of traits:
measurements of traits of individuals, 
and species-level traits.

Concretely, we have the following information about a number of whales:
species, sex,
and the volume and shapes of the right and left pelvis and (smallest) rib bones.
(Here ``shape'' is a high-dimensional measurement;
we ignore the details, assuming only we have a sensible method for measuring shape differences,
after removing size differences.)
For each species, we have
adult body length and
breeding male testes volume.
(Here I say ``we have''; but note that actually some values are missing,
a fact we will have to deal with.)

Furthermore, we assume that it is sensible to quantify the strength of prezygotic sexual competition in each species,
and hypothesize that the degree of competition affects testes volume, pelvic bone size, and speed of change of pelvic bone shape.
Not all forms of prezygotic sexual competition are expected to work in this way;
here we mean only the forms that are.

In summary, the variables we observe are, for species (or, node in the tree) $i$:
\begin{gather*}
    T_i = \text{(testes size)} .
\end{gather*}
Since the strength of sexual competition affects the evolution of other traits,
and is unobserved in the data,
it is better to associate a value of $C$ for each branch, rather than each taxon.
So, for a branch $e$ in the tree, define
\begin{gather*}
    C_{e} = \text{( strength of sexual competition )}  .
\end{gather*}
We will take $C$ to be a real number, with smaller (perhaps negative) values of $C$ denoting less competition.
For whale $j$ in species $i$, we have bones from both sides; so for $k \in \{\text{left},\text{right}\}$:
\begin{gather*}
    G_{ij} = \text{( sex )} \qquad  L_{ij} = \text{( body length )} \\
    V^P_{ijk} = \text{( pelvic volume )} \qquad S^P_{ijk} = \text{( rib shape )} \\
    V^R_{ijk} = \text{( rib volume )} \qquad S^R_{ijk} = \text{( rib shape )} 
\end{gather*}
Note this includes both observed and unobserved variables.
Furthermore, write, for instance $V^P_i$ for the mean pelvic volume in species $i$.

First we describe how the species-level variables evolve on the tree,
namely, across a single branch.
Write $L_r$ for the mean body length at the base of the branch, $L_t$ at the tip,
and likewise for other variables;
also let $\tau$ be the length of this branch.
We have defined competition to take values on the branches, not the nodes;
so first let $C_e$ denote the value of $C$ on this branch, 
and assume this is related to the value on the parent branch $e'$ by
\begin{align}
    C_e &= C_{e'} + Z^C \\
    Z^C &\sim N(0,\tau) .
\end{align}
Given the values at the base, the model is as follows:
\begin{align}
    L_t &= L_r \exp( Z^L ) \\
    T_t &= T_r \exp( \delta Z^L + \beta_T Z^C ) \\
    V^R_t &= V^R_r \exp( \delta Z^L + U^R ) \\
    V^P_t &= V^P_r \exp( \delta Z^L + U^P + \beta_P Z^C ) \\
    S^R_t &= S^R_r + Y^R \\
    S^P_t &= S^P_r + Y^P \\
    Z^L &\sim N(0,\sigma^2_L \tau) \\
    U^R &\sim N(0,\sigma^2_R \tau) \\
    U^P &\sim N(0,\sigma^2_P \tau) \\
    Y^R &\sim N(0,\sigma^2_S \tau I_{k_S}) \\
    Y^P &\sim N(0,(\sigma^2_S +\gamma_P C_e) \tau I_{k_S}) .
\end{align}
In words, the model is thus:
Competition, averaged over the branch, evolves by random (Gaussian) increments.
Length evolves multiplicitavely,
so log length evolves as Brownian motion on the tree,
with increments given by $Z^L$.
A multiplicative change $a$ in length changes volumes by a factor of $a^\delta$ (so, $e^{\delta Z^L}$);
in addition, there is independent evolution of each volume ($U$ terms).
Furthermore, changes in strength of competition affect the volume of the testes and pelvic bones.
Shape evolves as a $k_S$-dimensional Brownian motion (so changes are chi-squared),
and the speed of pelvic shape depends on the strength of competition.
In other words, if we let
$(\zeta_1, \ldots, \zeta_4)$ be independent N(0,1),
then we can sample across the edge by
\begin{align}
\begin{bmatrix}
    \log L_t \\ \log T_t \\ \log V^R_t \\ \log V^P_t 
\end{bmatrix}
=
\begin{bmatrix}
    \log L_r \\ \log T_r \\ \log V^R_r \\ \log V^P_r 
\end{bmatrix}
+
\begin{bmatrix}
    \sigma_L \sqrt{\tau} &  0  & 0  & 0  \\
    \sigma_L \sqrt{\delta \tau}  &  \beta_T \sqrt{\tau}  & 0  &   0 \\
    \sigma_L \sqrt{\delta \tau}  &  0        & \sigma_R \sqrt{\tau}  &   0 \\
    \sigma_L \sqrt{\delta \tau}  &  \beta_P \sqrt{\tau}  & 0  &   \sigma_P\sqrt{\tau}  
\end{bmatrix}
\begin{bmatrix}
    Z_1 \\ Z_2 \\ Z_3 \\ Z_4
\end{bmatrix}
\quad \begin{matrix}
    \leftarrow Z^L \\
    \leftarrow Z^C \\
    \leftarrow U^R \\
    \leftarrow U^P 
\end{matrix}
\end{align}


Given the species-level values,
the data are then
\begin{align}
    L_{ij} &=  L_i \exp( W^L_{ij} ) \\
    V^R_{ijk} &=  V^R_i \exp( \delta W^L_{ij} + W^R_{ij} + (-1)^k Z^R_{ij} ) \\
    V^P_{ijk} &=  V^P_i \exp( \delta W^L_{ij} + W^P_{ij} + (-1)^k Z^P_{ij} ) \\
    S^R_{ijk} &= S^R_i + Y^R_{ijk} \\
    S^P_{ijk} &= S^P_i + Y^P_{ijk} \\
    W^L_{ij} &\sim N(0,\zeta^2_L) \\
    W^R_{ij} &\sim N(0,\zeta^2_{R}) \\
    W^P_{ij} &\sim N(0,\zeta^2_{P}) \\
    Z^R_{ij} &\sim N(0,\omega^2_{R}) \\
    Z^P_{ij} &\sim N(0,\omega^2_{P}) \\
    Y^R_{ijk} &\sim N(0,\xi^2_R I_{k_S}) \\
    Y^P_{ijk} &\sim N(0,\xi^2_P I_{k_S}) \\
\end{align}
Here $W^L$ fits the unobserved length of the whale in question,
while $W^R$, $W^P$, and the shape differences can be thought of as measurement or developmental noise.
Similar to before, if we let $W_1, W_2, \ldots$ be independent N(0,1), 
then we can sample the individual observations of individual $j$, given the species, 
\begin{align}
\begin{bmatrix}
    \log L_{ij} \\
    \log V_{ij1}^R \\ \log V_{ij2}^R  \\
    \log V_{ij1}^P \\ \log V_{ij2}^P  
\end{bmatrix}
=
\begin{bmatrix}
    \log L_{i} \\
    \log V_{i}^R \\ \log V_{i}^R  \\
    \log V_{i}^P \\ \log V_{i}^P  
\end{bmatrix}
+
\begin{bmatrix}
    \zeta_L  &   0 & 0  & 0  & 0 \\
    \sqrt{\delta} \zeta_L  &  \zeta_R  &  \omega_R  & 0 & 0 \\ 
    \sqrt{\delta} \zeta_L  & \zeta_R   & - \omega_R & 0 & 0\\ 
    \sqrt{\delta} \zeta_L  & 0 & 0 &  \zeta_P  & \omega_P  \\ 
    \sqrt{\delta} \zeta_L  & 0 & 0 &  \zeta_P  & - \omega_P
\end{bmatrix}
\begin{bmatrix}
W_1 \\W_2 \\W_3 \\W_4 \\ W_5  \\ 
\end{bmatrix}
\quad \begin{matrix}
    \leftarrow W^L_{ij}  \\
    \leftarrow W^R_{ij}  \\
    \leftarrow Z^R_{ij}  \\
    \leftarrow W^P_{ij} \\ 
    \leftarrow Z^P_{ij} 
\end{matrix}
\end{align}
and different individuals are independent, given the species values.

Parameters, excluding those involving shape, are thus:
\begin{align}
    \theta_S = (
    \sigma_L,
    \beta_T,
    \beta_P,
    \sigma_R,
    \sigma_P,
    \zeta_L,
    \zeta_R,
    \omega_R,
    \zeta_P,
    \omega_P
    ) ,
\end{align}
and those involving shape are:
\begin{align}
    \theta_I = (
    \sigma^2_S,
    \gamma_P,
    k_S,
    \xi^2_R,
    \xi^2_P
    )
\end{align}
We only actually care about $\beta_T$, $\beta_P$, and $\gamma_P$.
We can furthermore add additional $\beta_R$ and $\gamma_R$ terms, which we expect to be zero.

\section{Implementation}

First extend the phylogenetic tree to include each individual sample, by adding a ``twig'' for each sample
to the tips that correspond to the species.
Then the above describes the changes across each branch, 
for $L$, $V$, and $S$,
as well as changes across all branches except the twigs for $T$,
and changes between branches, except the twigs, for $C$.
With the exception of shape changes, everything, appropriately transformed, 
is multivariate normal.
This suggests first fitting the model without $S$;
then using the posterior on $C$ to fit evolution of $S$.

To implement this, define the data vector, indexed by nodes $t$ in the tree, to be
\begin{align}
    X^L_t &= \log L_t \\
    X^T_t &= \log T_t \\
    X^P_t &= \log V^P_t \\
    X^R_t &= \log V^R_t .
\end{align}
For species, nodes ($t=i$) we observe
\begin{align}
    X^L_i &= \log L_i \\
    X^T_i &= \log T_i ,
\end{align}
with $X^P_i$ and $X^R_i$ unobserved.
and for individuals we observe ($t={ijk}$):
\begin{align}
    X^L_{ijk} &= \log L_{ij} \\
    X^P_{ijk} &= \log V^P_{ijk} \\
    X^R_{ijk} &= \log V^R_{ijk} ,
\end{align}
with $X^L_{ijk}$ and $X^T_{ijk}$ unobserved.
Furthermore, associate to each note the value of $C$ on the edge leading to it,
with $C_\rho=0$ at the root, and $C_{ijk}=0$ on the twigs.

Let $X_\rho$ denote the values at the root, 
and for each edge $e$ from node $s$ to node $t$, let $dX_e = X_t - X_s$.
Then if $\rho \to t$ is the set of edges leading from $\rho$ to $t$, 
\begin{align}
    X_t = X_\rho + \sum_{e \in \rho \to t} dX_e .
\end{align}
Each $dX_e$ is independent; we assume the values at the root are also;
above, we have described the covariance matrix for each $dX_e$:
we have written $dX_e = A Z$, with $Z$ independent N(0,1);
hence 
\begin{align}
    \cov[dX_e] = A A^T .
\end{align}
Also, $A$ is one of two forms, depending on if it is an internal edge (on the species tree)
or if it on a tip (leading to a sample): 
if we set edge lengths on the tips to be 1,
then either $A = \sqrt{\tau_E} \Sigma_I$ or $A=\sqrt{\tau_E} \Sigma_T$.

Furthermore, notice that 
\begin{align}
  \cov[X_s,X_t] &= \sum_{e \in \rho \to t \cap \rho \to s } \cov[dX_e] \\
    &= \sum_{e \in \rho \to t \cap \rho \to s \cap I } \cov[dX_e] 
      + \sum_{e \in \rho \to t \cap \rho \to s \cap T } \cov[dX_e] \\
    &= \left( \sum_{e \in \rho \to t \cap \rho \to s \cap I } \tau_e \right) \Sigma_I
      + \left( \sum_{e \in \rho \to t \cap \rho \to s \cap T } \tau_e \right) \Sigma_T 
\end{align}

In other words, as in Revell \& Collar 2009, the covariance matrix between species nodes is $C \otimes \Sigma_I$,
where $C_{ij}$ is the total time in the tree since the common ancestor of $i$ and $j$,
and $\otimes$ is the Kronecker product.
Those at the tips all have the same for $\Sigma_T$.
So, if we extend the matrix $C$ to be indexed by samples, 
so that $C_{s,t}$ gives the time in the tree between the species that $s$ and $t$ belong to;
and let $T_{s,t} = 1$ if $s$ and $t$ are the same species, and 0 otherwise;
then the full covariance matrix is
\begin{align}
    \Sigma = C \otimes \Sigma_I + T \otimes \Sigma_T .
\end{align}

We also want to subtract off the mean across traits,
since $X$ still depends on the value at the root.

If we normalize the $X$ by subtracting off a weighted mean across species: 
\begin{align}
\tilde X_t = X_t - \sum_t w_t X_t,
\end{align}
with $\sum_t w_t = 1$,
then $\tilde X$ does not depend on $X_\rho$:
if we define 
\begin{align}
    w_e = \sum_{t : e \in \rho \to t} w_t
\end{align}
then
\begin{align}
    \tilde X_t = \sum_{e \in \rho \to t} dX_e - \sum_e w_e dX_e,
\end{align}
which doesn't include $X_\rho$.

The covariance matrix is now easy to compute:
if $B^T B = \Sigma$ is the Cholesky decomposition of $\Sigma$, then we can write $X = B^T Z$,
with $Z$ independent N(0,1);
and if $W$ is the matrix whose rows are given by $w$,
\begin{align}
  \tilde X = (I-W) B^T Z,
\end{align}
and hence
\begin{align}
  \cov[\tilde X, \tilde X] := \tilde \Sigma = (I-W)B^T B (I-W^T) = (I-W) \Sigma (I-W^T) .
\end{align}
To compute the likelihood function, we let $A^T A = \tilde \Sigma$ be the Cholesky decomposition of $\tilde \Sigma$;
so that $\tilde \Sigma^{-1} = (A^T A)^{-1} = A^{-1} (A^T)^{-1}$, and hence
\begin{align}
  x^T \tilde \Sigma^{-1} x = \| (A^T)^{-1} x \|^2 ,
\end{align}
and since $A$ is upper triangular, $A^{-1} x$ can be found efficiently, e.g.\ with \texttt{forwardsolve(t(A),x)} in R.


\section{Inference}

So, we've reduced the model to a big multivariate normal distribution,
where all the parameters are variances.
Now we can put priors on the variances and MCMC to get the posterior distribution.


\section{The likelihood computation}

If $X$ is an $n$-dimensional multivariate Gaussian with mean 0 and covariance matrix $\Sigma$ that depends on parameters $\theta$,
then the likelihood function is
\begin{align}
  \mathcal{L}(x) = \frac{1}{ (2 \pi)^{n/2} \sqrt{ \det \Sigma } } \exp \left( - \frac{1}{2} x^T \Sigma^{-1} x \right)
\end{align}
and so the negative log-likelihood function is
\begin{align}
  L(x) = \frac{1}{2} \log \left( \det \Sigma \right) + \frac{1}{2} x^T \Sigma^{-1} x  .
\end{align}
If $\Sigma = A^T A$, with $A$ upper triangular, then $\Sigma^{-1} = (A^T A)^{-1} = A^{-1} (A^T)^{-1}$, and so
if $z$ solves $A^T z = x$, then
\begin{align}
  x^T \Sigma^{-1} x &= x^T A^{-1} (A^T)^{-1} x \\
  &= ( (A^T)^{-1} x )^T ( (A^T)^{-1} x ) \\
  &= \| (A^T)^{-1} x \|^2 \\
  &= \|z\|^2  .
\end{align}
Also, 
\begin{align}
  \det \Sigma = (\det A )^2 = \left( \prod_{i=1}^n A_{ii} \right)^2 .
\end{align}


\section{Shape differences}

Recall above that we can write the shape difference between two particular nodes, 
say $s$ and $t$ by
\begin{align*}
    S_t - S_s = \sum_{e \in s \to t} Y_e ,
\end{align*}
with the $Y_e$ a vector of independent Gaussians with variance we define to be $\tau'_e$, 
which is either $(\sigma^2_S + \gamma_P C_e) \tau_e$ on internal edges, or $\xi^2_P$ on the tips.
The parameters only enter through these rescaled branch lengths.
Now 
\begin{align}
  \|S_t - S_s\|^2 &= \sum_k \| \sum_{e \in s \to t} Y_{e,k} \|^2 , \quad \text{and} \\
  \sum_{e \in s \to t} Y_{e,k} & \sim N(0, \sum_{e \in s \to t} \tau'_e ), \quad \text{so} \\
  \frac{ \sum_{e \in s \to t} Y_{e,k} }{ \sqrt{ \sum_{e \in s \to t} \tau'_e } } &\sim N(0,1), \quad \text{and hence} \\
  \frac{ \|S_t - S_s\|^2 }{  \sum_{e \in s \to t} \tau'_e } &\sim \chi^2_{k_s} .
\end{align}
For large $k_s$ we can approximate the chi-squared as Gaussian,
with parameters
\begin{align*}
    \E[ \|S_t - S_s\|^2 ] &= k_s \sum_{e \in s \to t} \tau'_e \\
    \var[ \|S_t - S_s\|^2 ] &= 2 k_s \left( \sum_{e \in s \to t} \tau'_e \right)^2 
\end{align*}
Now, the covariance: note that if $Z$ is independent of $X$ and $Y$, and has mean 0, then
\begin{align}
    \cov[X,(Y+Z)^2] &= \cov[X,Y^2] + 2 \cov[X,YZ] + \cov[X, Z^2] \\
    &= \cov[X,Y^2] + 2 \E[Z] \cov[X,Y] + \cov[X,Z^2] \\
    &= \cov[X,Y^2] ,
\end{align}
and hence, by iteratively removing edges not in $(u \to t) \cap (u \to s)$,
since $\E[Y_e]=0$,
\begin{align}
    \cov[ \|S_t - S_u\|^2, \|S_s - S_u\|^2 ] &= \sum_{k=1}^{k_s} \cov\left[ (S_t-S_u)_k^2, (S_s-S_u)_k^2 \right] \\
          &= \sum_{k=1}^{k_s} \var\left[ \left( \sum_{e \in (u \to t) \cap (u \to s)} Y_{e,k} \right)^2 \right] \\
                                             &= \var\left[ \|S_{t \wedge s} - S_u\|^2 \right] \\
                                             &= 2 k_s \left( \sum_{e \in (u \to t) \cap (s \to t) } \tau'_e \right)^2 
\end{align}
where $t \wedge s$ is the node where the paths $u \to t$ and  $u \to s$ diverge.

So, we're back to a multivariate Gaussian distribution;
in this case the parameters $k_s$, $\sigma^2_S$, $\gamma_P$, and $\xi^2_P$ are used to construct $\tau'_e$,
which then is used to make the covariance matrix between the $\|S_t - S_u\|^2$.



\end{document}





If we normalize the $X$ by subtracting off a weighted mean across species: 
\begin{align}
\tilde X_t = X_t - \sum_t w_t X_t,
\end{align}
with $\sum_t w_t = 1$,
then $\tilde X$ does not depend on $X_\rho$:
if we define 
\begin{align}
    w_e = \sum_{t : e \in \rho \to t} w_t
\end{align}
then
\begin{align}
    \tilde X_t = \sum_{e \in \rho \to t} dX_e - \sum_e w_e dX_e,
\end{align}
and if each branch's covariance is a constant multiplied by a common matrix $\Sigma$, $\cov[dX_e] = \tau_e \Sigma$, 
then the covariance between two observations is just
\begin{align}
    \cov[\tilde X_s,\tilde X_t] &= \sum_e \left( \mathbf{1}_{e \in \rho \to s} - w_e \right) \left( \mathbf{1}_{e \in \rho \to t} - w_e \right) \cov[dX_e] \\
                                &= \left( \sum_{e \in \rho \to t \cap \rho \to s} \tau_e - \sum_{e \in \rho \to t} w_e \tau_e - \sum_{e \in \rho \to s} w_e \tau_e + \sum_e w_e^2 \tau_e \right) \Sigma \\
                                &= \left( \sum_{e \in \rho \to t \cap \rho \to s} (1-2w_e) \tau_e - \sum_{e \in s \to t} w_e \tau_e + \sum_e w_e^2 \tau_e \right) \Sigma .
\end{align}
